Refining character relationships using embeddings of textual units

### 2022-09-01 article/paper1.tex
write article

Distant reading tools allow researchers, from various fields, to quickly gain knowledge on textual corpora without actually reading them. Purposes of these methods are various, but can be mainly categorized into two groups: in the first case, these methods are used in order to tag, classify, or summary large quantities of texts, in order to quickly structure information or to deliver a speech over the whole studied corpus. Methods in this case rely heavily on Big Data and make an extensive use of Machine Learning algorithms. In the second case, researchers use these methods to underline hidden structures in a particular text, helping them to refine their understanding of it and reinforce stated hypotheses. Methods in this setting can also rely on Machine Learning, but must typically be build with more caution and attention to details: corpus are smaller, analyses are closer to the work, and methods must be more transparent in order to appropriately interpret results. \\
Automatic extraction and analysis of \emph{character networks} from literacy works typically belong in the latter group. These methods aim at representing various interactions occurring between fictional characters found in a textual narrative with a graph, thus showing explicitly the hidden structure of character relationships constructed by the author. This structure might allow to find hidden pattern within book, which can highlight a particular genre or author style and help to understand a part of the "flavor" given by the author to the text.

\section{Methods}

When building character networks from a textual narrative, the most widespread method consists in dividing the studied work into $n$ textual units $u_1, \ldots, u_n$, which can be, e.g., sentences, paragraphs, or chapters, and counting characters co-occurrences in these units. Usually, the text constituting these units is discarded and the resulting network displays edges which roughly represent an aggregated number of interactions between characters. However, by doing so, the aggregation occurs on various type of interactions and will give little information about the type of relationship which exist between characters. In this paper, we propose a data organization, leading to various type of analyses, which permits the use of the text contained in the unit in order to characterize relationship

@@

A textual narrative divided in $n$ textual units $u_1, \ldots, u_n$ can be represented in a $n \times (p+1)$ table $N$, where $p$ is the number of characters found in the narrative. Each line represent a textual unit, the first column is the text composing this unit, and the remaining $p$ columns contains the number of character

>>>>

Distant reading tools allow researchers, from various fields, to quickly gain knowledge on textual corpora without actually reading them. Purposes of these methods are various, but can be mainly categorized into two groups: in the first case, these methods are used in order to tag, classify, or summary large quantities of texts, in order to quickly structure information or to deliver a speech over the whole studied corpus. Methods in this case rely heavily on Big Data and make an extensive use of Machine Learning algorithms. In the second case, researchers use these methods to underline hidden structures in a particular text, helping them to refine their understanding of it and reinforce stated hypotheses. Methods in this setting can also rely on Machine Learning, but must typically be build with more caution and attention to details: corpus are smaller, analyses are closer to the work, and methods must be more transparent in order to appropriately interpret results. \\
Automatic extraction and analysis of \emph{character networks} from literacy works typically belong in the latter group. These methods aim at representing various interactions occurring between fictional characters found in a textual narrative with a graph, thus showing explicitly the hidden structure of character relationships constructed by the author. This structure might allow to find hidden patterns within a book, which can highlight a particular genre or style.

\section{Methodology}

When building character networks from a textual narrative, the most widespread method consists in dividing the studied work into $n$ textual units $u_1, \ldots, u_n$, which can be, e.g., sentences, paragraphs, or chapters, and then counting characters co-occurrences in these units. Usually, the text constituting these units is discarded and the resulting network displays edges which roughly represent an aggregated number of interactions between characters. However, by doing so, the aggregation occurs on various type of interactions and will give little information about the type of relationship which exist between characters. In this paper, we propose a data organization leading to various methods which takes into consideration the text contained in the unit, helping to refine understanding of characters and their relationships.

@@

In this article, the data representation for a textual narrative, divided in $n$ textual units $u_1, \ldots, u_n$, is made through two tables. The first one is well known in the field of textual analysis, and consists in the $n \times v$ contingency table $\mathbf{N}$, as represented by Table \ref{cont_table}, where $v$ is the vocabulary size. In this table, each row represents an unit, each column a word, and cells $n_{ij}$ counts the number of times the word $i$ appears in the unit. Using this table typically denotes a \emph{Bag-of-Words} approach in our analyses. \\
\begin{table}[h]
	\scriptsize
	\begin{tabular}{|c||c|c|c|c|c|c|}
		\hline
		& aller & allumer & apercevoir & bas & bon & ... \\
		\hline
		\hline 
		$u_{100}$ & 23 & 2 & 6 & 11 & 6 & ... \\
		\hline
		$u_{7795}$ & 12 & 1 & 0 & 3 & 9 & ... \\
		\hline
		$u_{7796}$ & 10 & 0 & 5 & 1 & 5 & ... \\
		\hline
		$u_{7797}$ & 0 & 0 & 1 & 0 & 0 & ... \\
		\hline
	\end{tabular}
	\label{cont_table}
	\caption{A snippet of the contingency table $\mathbf{N}$ extracted from \emph{les Misérables}. Rows are chapters, columns are words in the vocabulary, and cell $n_{ij}$ counts the number of time word $j$ appear in chapter $i$.}
\end{table}
The second table, denoted by $\mathbf{O}$, has a size of $n \times p$ where $p$ is the number of \emph{character-based objects} found in the narrative and cells $o_{ij}$ counts occurrences of object $j$ in unit $i$. A character-based object can be loosely defined in order to be flexible for various types of narrative or analyses, but can be roughly seen as a countable recurring narrative entity, containing one or more characters. For example, it can be a character occurrence, a character co-occurrence, an oriented character interaction (e.g. a dialog), or even a particular recurring event containing multiple characters (e.g. a lecture). In this article, we mostly consider character occurrences and pair of characters co-occurrences, as shown in Table \ref{occ_table}.

\begin{table}[h]
	\scriptsize
	\begin{tabular}{|c||c|c|c|c|}
		\hline
		& Cosette & Thénardier & Valjean & ... \\
		\hline
		\hline 
		$u_{101}$ & 66 & 78 & 0 & ... \\
		\hline
		$u_{102}$ & 21 & 26 & 0 & ... \\
		\hline
		$u_{103}$ & 12 & 25 & 0 & ... \\
		\hline
		$u_{104}$ & 12 & 0 & 5 & ... \\
		\hline
	\end{tabular}
	\begin{tabular}{|c||c|c|c|c|}
		\hline
		& ... & Cosette-Thénardier & Cosette-Valjean & ... \\
		\hline
		\hline 
		$u_{101}$  	& ... & 66 & 0 &  ... \\
		\hline
		$u_{102}$ 	& ... & 21 & 0 & ... \\
		\hline
		$u_{103}$  	& ... & 12 & 0 & ... \\
		\hline
		$u_{104}$ 	& ... & 0 & 5 & ... \\
		\hline
	\end{tabular}
	\label{occ_table}
	\caption{A snippet of the character-based objects table $\mathbf{O}$ extracted from \emph{les Misérables}. Rows are chapters, columns are character occurrences (top) and character co-occurrences (bottom), and cell $o_{ij}$ counts the number of times object $j$ appear in chapter $i$. Character co-occurrences are computed here as $o_{ij} = \min(o_{ik}, o_{il})$, where $\{k, l\}$ are characters constituting $j$.}
\end{table}

This data organization already gives an orientation to the subsequent analyses and should be kept in mind of the practitioner. Textual unit are now considered as \emph{individuals} (in the statistical terminology), defined by their \emph{variables} contained in the different columns of both tables. Moreover, subsequent analyses are oriented in searching how the object table $\mathbf{O}$ has an influence over the contingency table $\mathbf{N}$, i.e. searching which words are over-represented or under-represented knowing the character-based objects in the unit. While authors use characters in order to build the narrative, we, to a certain extent, work backward: we are searching how character appearances and interactions in the textual unit act on her/his choice of words. If extraction methods permit it, a practitioner should include all character-based objects which she/he desire to study. Here, for example, the choice to include character co-occurrences along with occurrences is motivated by the fact that we are interested in studying character relationships. A character co-occurrence can roughly be seen as an interaction between two characters, and an interaction between two characters (or more, but higher order interactions are outside the scope of this article) should be considered as an object of its own: this interaction is not necessarily the sum of its parts and gives a particular flavor to the unit. \\
This data organization also highlight the importance of choosing a proper size for the units. These units should be large enough to contain enough words in order the properly capture the textual specificity of each unit, but not too large, as each unit should ideally capture particularities about one of the character-based objects. Unfortunately, it is impossible to define an ideal size for all types of analysis, and this size should be balanced regarding the level of analysis, text size, selected character-based objects, and previous knowledge of the studied work. \\
The use of a contingency table to represent the textual resource present in the units denotes a \emph{Bag-of-Words} approach. Using this approach loses the information relative to the order of words in the units, but permits to transform a chain of characters, improper to statistical analyses, into a contingency table, a well studied mathematical object which permits the use of various kind of statistical methods. The next section shows two methods of analysis based on this table. \\

\subsection{Embedding of textual units}

Various methods can be performed on the contingency table $\mathbf{N}$ in order to extract information from it, such as Sentiment Analysis (REF), Textometry (REF), or even Deep Learning methods. Here, we make to choice to extract a lower dimensional, numeric representation of each units, in other words, an \emph{embedding of textual units}. In section (REF) These embeddings will help to see how resulting vectors depends on character occurrences and co-occurrences. Various kind embeddings can be performed, and we will focus here on two methods. Section (REF) describes \emph{Correspondence Analysis (CA)} and section (REF) focuses on \emph{Pre-trained, static, word vectors (WV)}.

\subsubsection{Correspondence Analysis}

Using Correspondence Analysis in order to analyse textual resources has a long tradition (REF). It has the advantage to provide an embedding space, the factorial map, where units are placed alongside words, allowing to analyse or represent the different units in terms of word specificities. Moreover, units and words placement in the embedding space have a direct interpretation in terms of chi2 distance and profiles, which is desirable when interpreting results. \\
By performing the Correspondence Analysis, we get $n$ vectors $\mathbf{x}_1, \ldots, \mathbf{x}_n$ corresponding to units (rows) of table $\mathbf{N}$ and $v$ vectors $\mathbf{w}_1, \ldots, \mathbf{w}_v$ corresponding to words (columns). Each of these vectors has a size of $\min(n, v) - 1$, which will generally be $n - 1$.  (MORE EXPLANATION ON CA ? SEE MQ IV).  \\
Note that vectors $\mathbf{x}_1, \ldots, \mathbf{x}_n$  obtained from CA reflect textual unit profile (in terms of words)  regarding the mean profile. This mean that this analysis is in fact \emph{contrastive}: we highlight unit variations among the studied text. This could therefore becomes problematic if we would like to build our dataset on multiple work: units belonging to the same text have a great chance to be close from each other, and the displayed variance might be largely composed by differences of style between text.

\subsubsection{Pre-trained word vector}

Pre-trained word vectors, based on Word2Vec (REF), Glove (REF), or Fasttext (REF), have recently received great attention from various fields (REFS). The generally originate from a training of very large corpora, such as Wikipedia (REF) or Common Crawl (REF) and results in a embedding containing a large quantities of word vectors.

\subsubsection{Character and character pairs embeddings}
